\chapter{Introduction}\label{chap:introduction}


\textit{Why should one pay attention to secondary financial markets, or namely one of those markets, the stock market?} \citet{bondetal2012}, motivated to verify the impact of the secondary financial market on the real economy, argue that the role of market prices' information is critical to real decision makers (e.g., firm managers, customers, or employees). There are different ways to look at this role. The first instant can be when the decision makers decide based on stock price, e.g., credit rating. Even if one does not make a decision directly, one is still motivated to pay attention to stock price, for example, when an employee's compensation is affected by the company's share price. There are also developed theoretical models that contain the "feedback effect" between the financial market and the real economy.


As stock prices are informative and thus important, many efforts have been made in the literature to investigate their behavior in both theoretical and empirical territories. Looking back through the history of the trading world, from the times of crowded trading floors to the modern days with algorithmic trading, financial data structure transforms from lower frequencies (e.g., daily, weekly, monthly, or annually) to the (ultra-) high frequencies, such as milliseconds or nanoseconds. The high frequency financial data owns some specific characteristics, including the discreteness of transaction price and irregular time between trades \citep{tsay2013}. The discreteness of transaction price challenges the traditional assumption of a continuous probability distribution. \citet{hausman1992} address this problem by introducing an ordered probit model that considers price discreteness. Moreover, their model also captures the effect of market microstructure elements on price, verifying the "information-effect" theory by \citet{easleyohara1987}. One further application of their model, which interests many traders, is to predict the next price movement. Since then, although the so-called "low-latency" trading world continues to evolve and imposes many challenges, research directions (e.g., \citet{ohara2015}), as well as newer machine learning models are providing promising results, the ordered probit model's ability to serve both market microstructure and forecasting questions is worth a kind of "test-of-time" study.

Hence, this thesis would revisit the \citet{hausman1992}'s ordered probit model, attempting to examine the following objectives:
\begin{itemize}
    \item \textit{Does the information of the sequence of previous price changes affect the next price change? Does the trading volume of the previous trade affect the next price change?}
    \item and \textit{How effective is our model in forecasting the stock price movement?}
\end{itemize}

{\noindent {By answering these questions, we hope to learn more about how the financial market microstructure works and the effectiveness of the ordered probit model in our settings.}}


The remainder of the thesis is organized as follows. Chapter \ref{chap:relatedwork} provides an overview of the related literature, specifically in terms of predicting stock price movement. Chapter \ref{chap:model} describes our simplified ordered probit model and its specification. Chapter \ref{chap:mainresults} discusses the empirical results of the main study. Chapter \ref{chap:extension} presents the further applications of our model to two special periods: the Covid-19 period in 2020, and the financial crisis in 2008. Chapter \ref{chap:conclusion} reviews the limitations and challenges of this thesis and concludes our findings.